Machine Learning Methods for Discourse Segmentation of Communications in E-Mail Based Behavioral Interventions

### 2018-05-24 paper-amia-2018/manuscript.tex
update manuscript

\author{Mehedi Hasan, BS$^{1}$\footnote[1]{Authors provided equal contribution. \label{footnote1}}, Alexander Kotov, PhD$^{1}$\textsuperscript{\ref{footnote1}}, Sylvie Naar, PhD$^{2}$, Gwen L. Alexander, PhD$^{3}$, April Idalski Carcone, PhD$^{4}$}

@@

\noindent{\bf Abstract}
\textit{Communication science approaches to developing effective behavior interventions, such as motivational interviewing (MI), are limited by traditionally manual qualitative coding of communication exchanges, which is a very resource-intensive and time-consuming process. This study focuses on the analysis of e-Coaching sessions, behavior interventions that are delivered via email and grounded in the principles of MI. A critical step towards automated annotation of e-Coaching communication exchanges is segmentation of emails into textual fragments that correspond to MI behaviors. In this work, we formulate this task as a classification problem and propose lexical, punctuation and topic features to address it. We experimented both with traditional machine learning methods, such as Support Vector Machine (SVM), Naive Bayes and K-Nearest Neighbor (KNN) classifiers and recurrent neural networks (RNNs). Results indicate that SVM outperformed KNN and RNNs achieving 0.990 macro F1-score overall, and 0.848 and 0.995 macro F1-score for detecting ``new segment'' and ``same segment'' classes, respectively.}

@@

Qualitative coding of motivational interviews with pre-defined codes has been traditionally performed manually by trained annotators, which is a tedious and resource-intensive process that involves several iterations of reading, comprehension and interpretation of interview transcripts. Rapidly developing computational technologies, specifically, machine learning methods, offer a unique opportunity to accelerate this process. In particular, machine learning methods have been successfully applied to a variety of analytical tasks involving textual data, such as classification\cite{nigam2000text} and sentiment analysis\cite{wang2012baselines}. In our previous work, we examined the utility of machine learning methods for automated annotation \cite{hasan2016study,kotov2015interpretable} and analysis \cite{hasan2018predicting} of in-person MI sessions. Specifically, we demonstrated that machine learning methods can be utilized for annotation of MI transcripts according to a simple communication code scheme with the accuracy comparable to human coders\cite{hasan2016study}. Experimental data utilized in these studies, however, were prepared by transcribing audio conversations, which were clearly segmented into utterances by a counselor, a patient, and, in some cases, a caregiver.

@@

The goal of this research study is to assess the applicability of machine learning methods for automated segmentation of e-Coaching emails into textual fragments corresponding to individual behaviors, which is the first step of the coding process of e-Coaching communications. In particular, we introduced lexical, topic and punctuation features and experimented with both traditional supervised machine learning methods, such as Support Vector Machine (SVM), Naive Bayes (NB) and K-Nearest Neighbor (KNN) classifiers, and deep learning methods, such as Long Short Term Memory (LSTM) and Gated Recurrent Unit (GRU), to find the best performing method and feature combination.

Relevant previous work in the biomedical domain primarily focused on segmentation of text into sections and headers\cite{apostolova2009automatic,denny2009evaluation,tepper2012statistical,cho2002text} or sentence boundary detection\cite{griffis2016quantitative,kreuzthaler2015detection,treviso2016sentence}. Apostolova et al.\cite{apostolova2009automatic} applied SVM along with word-vector cosine similarity metric combined with several heuristics to segment clinical reports into sections, such as demographics, history, procedure, finding and impression. After identification of each line in the document, Tepper et al. \cite{tepper2012statistical} trained Maximum Entropy models for section classification. Denny et al.\cite{denny2009evaluation} proposed a SecTag algorithm, which combined natural language processing techniques, terminology-based rules, and Naive Bayes classifier to identify the sections and headers that achieved 99\% recall with 95.6\% precision. On the other hand, SVM based on prosodic and part of speech features \cite{kreuzthaler2015detection} and recurrent convolutional neural networks using word embeddings \cite{griffis2016quantitative} were utilized for detecting sentence boundaries. Segmentation of e-Coaching emails is different from traditional shallow discourse analysis of conversations \cite{galley2003discourse} in that the focus is on segmentation, rather than on determining the types of transitions between the utterances or assigning utterances to speakers.

@@

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from 49 e-coaching sessions, which include a total of 3,138 segmented and annotated MI behaviors. Each session represents an MI intervention delivered via email. During pre-processing, we removed all non-ASCII characters and applied stemming to normalize morphological variants of related concepts, such as ``eating'', ``eats'', and ``eat''. We formulate the segmentation task as a binary classification problem, as illustrated in Figure~\ref{fig:classifier} and experimented both with traditional machine learning methods, such as Support Vector Machine (SVM), Naive Bayes (NB) and K-Nearest Neighbor (KNN) classifiers and recurrent neural networks (RNNs), such as Long Short Term Memory (LSTM) and Gated Recurrent Unit (GRU). For NB, SVM and KNN models, e-Coaching email exchanges are partitioned into adjacent word pairs. Each pair is classified into either ``new segment'' or ``same segment'' class. The gold standard is manually segmented where adjacent word pairs are classified into ``new segment'' class. If all adjacent word pairs of a block of text are classified into ``same segment'' class, the entire block is treated as one textual fragment corresponding to a single MI behavior. In total, we obtained 95,421 word pairs, which include 3,138 ``new segment'' and 92,283 ``same segment'' instances. For RNNs, a block of text was taken as input sequence, such that one-hot encodings of each word or punctuation marks in a block were used as input into an RNN and binary labels (1 or 0) corresponding to ``new segment'' and ``same segment'' classification decision were considered as the output of RNN at each step. In the gold standard, words within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1.

@@

where $p(l)$ is the prior probabilities of label $l$ in the training set and $p(w)$ is the prior probability of word $w$ in the language model estimated from the training set using maximum likelihood. We use the distribution $p(l|w)$ as the topic feature vector, which shows how indicative each word $w$ is for a particular behavior code $l$. Topic features allow to abstract away from individual words and detect segment boundaries by indirectly capturing transitions between behavior codes when only observing a pair of words that are specific to different behavior codes. To find the optimal topic feature, we also derived the same topic feature $p(l|w)$ by using two recently proposed probabilistic generative latent variable models: Discriminative Labeled Latent Dirichlet Allocation (DL-LDA) and Latent Class Allocation (LCA)\cite{kotov2015interpretable}. Similar as L-LDA, DL-LDA-based topic feature is extracted by first deriving a label-specific language model $p(w|l)$ for each label $l$ from the word distributions $p(w|l,z)$ for topic $z$ specific to label $l$ by using the Eq.~\ref{eq:pwl} and \ref{eq:plw}. On the other hand, LCA directly provides label-specific multinomials $p(w|l)$, which can be used to derive topic feature by using Eq.~\ref{eq:plw}. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;', `-'\} between a pair of words, are also employed as a feature, since punctuation marks designate the boundary of a sentence, clause, and phrase and often also correspond to segment boundary.

@@

\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\label{eq:resetgru}
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\label{eq:candidategru}
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
In Eq.~\ref{eq:firstgru}-\ref{eq:lastgru}, $\sigma$ corresponds to sigmoid function and $\odot$ designates an element-wise product. The update gate $z_t$ and reset gate $r_t$ at time step $t$ are computed by the Eq.~(\ref{eq:firstgru}) and~(\ref{eq:resetgru}), where $W_z$, $W_r$, $W_h$, $U_z$, $U_r$, $U_h$ are the weight matrices and $b_z$, $b_h$ and $b_r$ are bias vectors. The activation $h_t$ of the GRU at time $t$ is a linear combination of previous activation $h_{t-1}$ and the candidate activation $\tilde h_t$, which is represented by Eq.~(\ref{eq:lastgru}) and~(\ref{eq:candidategru}). Our GRU model includes one hidden layer, output layer, and input layer. We reset our model state after feeding each input sequence where input was given as one-hot encoding of word vector. Since LSTM and GRU use one-hot vector as input, results for these models are reported only when lexical and punctuation features are used. We experimentally determined that the best performance is achieved when the number of hidden units is 25, batch size is 1, and Adam\cite{kingma2014adam} is used for optimization.

\subsection*{\textit{Evaluation metrics}}
We report standard metrics for experiments (precision, recall, and F1-measure) to evaluate the performance of binary classifiers\cite{aas1999text}. However, accuracy is not reported as a performance metric because accuracy is highly sensitive to the prior class probabilities and does not fully describe the actual difficulty of the decision problem for an unbalanced dataset. The results are reported based on 5 folds cross-validation and weighted macro-averaging over the folds. We also estimate the area under the receiving operating characteristics (ROC) curve\cite{kumar2011receiver} (AUC) metric due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets \cite{hu2015kernelized}.

\section*{Results}
Our experimental results have two important directions. First, results are reported with respect to ``new segment'' and ``same segment'' classes as well as their weighted average in Table~\ref{tab:result_base}. Second, classification performance of different machine learning methods are outlined in Tables~\ref{tab:result_boundary}, \ref{tab:result_not_boundary}, and \ref{tab:result_weighted_avg} when topic features are extracted with different probabilistic generative latent variable models.\\

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM, and KNN methods for identification of ``new segment'' and ``same segment'' classes as well as their weighted average. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_base}
  \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{new segment}} & \multicolumn{3}{|c|}{\textbf{same segment}}  & \multicolumn{3}{|c|}{\textbf{overall}} \\\cline{2-10}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1}\\ \hline    
 \multicolumn{10}{|c|}{lexical features only} \\ \hline   
 NB & 0.594 & 0.662 & 0.626 & 0.988 & 0.985 & 0.987 & 0.975 & 0.974 & 0.975  \\ \hline
 SVM & 0.762 & \textbf{0.673} & 0.715 & \textbf{0.989} & 0.993 & 0.991  & 0.981 & 0.982 & 0.982 \\ \hline
 KNN & \textbf{0.808} & 0.663 & \textbf{0.728} & \textbf{0.989} & \textbf{0.995} & \textbf{0.992} & \textbf{0.983} & \textbf{0.984} & \textbf{0.983} \\ \hline
 \multicolumn{10}{|c|}{lexical and puctuation features} \\ \hline
 LSTM & \textbf{0.800} & 0.646 & 0.714  & 0.993 & \textbf{0.994} & \textbf{0.994} & \textbf{0.986} & 0.983 & 0.984 \\ \hline
 GRU & \textbf{0.800} & \textbf{0.715} & \textbf{0.741} & \textbf{0.994} & \textbf{0.994} & \textbf{0.994} & \textbf{0.986} & \textbf{0.985} & \textbf{0.986} \\ \hline
  \end{tabular}
\end{table}

As follows from Table~\ref{tab:result_base}, KNN outperforms all other models in terms of precision and F1-measure achieving 0.808 precision with 0.728 F1-measure for new segment detection. KNN also shows superior performance in all performance metrics for ``same segment'' class and wighted average over ``new segment'' and ``same segment'' classes. NB demonstrates the lowest performance among all models in terms of all performance metrics. On the other hand, SVM has the highest recall of 0.673 when only lexical features are used to identify ``new segment''. Results indicate that performance of classifiers is remarkably higher for ``same segment'' class compared to ``new segment'' class, which is expected, since 96.71\% instances belong to ``same segment'' categories. For example, KNN achieves 22.40\%, 50.07\%, and 36.26\% higher precision, recall, and F1-measure, respectively, in same segment identification compared to new segment detection. When lexical features are used in combination with punctuation mark, GRU demonstrates the best performance in all cases showing 10.68\% higher recall and 3.78\%  higher F1-measure than LSTM for new segment identification, 0.1\% higher precision than LSTM for same segment identification and 0.2\% higher recall and 0.2\%  higher F1-measure than LSTM for overall classification. \\

@@

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM, and KNN methods with different topic model-based features for identification of ``new segment'' class in e-Coaching emails, when all features are used together. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_boundary}
  \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{L-LDA-based topic features}} & \multicolumn{3}{|c|}{\textbf{DL-LDA-based topic features}}  & \multicolumn{3}{|c|}{\textbf{LCA-based topic features}} \\\cline{2-10} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1}\\ \hline

@@

Table~\ref{tab:result_boundary} summarizes the performance of NB, SVM, KNN, and RNN models for detecting ``new segment'' class. We observed that performance is remarkably better for ``new segment'' class compared to new segment detection, which is expected, since 96.71\% instances belong to ``same segment'' class. GRU demonstrates the same result as KNN for precision and F1-measure, but not for recall. SVM shows moderate performance, achieving 0.981, 0.982, and 0.983 for precision, recall, and F1-measure, respectively, when lexical features are used. \\

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM, and KNN methods with different topic model-based features for identification of ``same segment'' class in e-Coaching emails, when all features are used together. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_not_boundary}
  \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{L-LDA-based topic features}} & \multicolumn{3}{|c|}{\textbf{DL-LDA-based topic features}}  & \multicolumn{3}{|c|}{\textbf{LCA-based topic features}} \\\cline{2-10} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1}\\ \hline

@@

As shown in Table~\ref{tab:result_not_boundary}, performance of classifiers is remarkably higher for ``same segment'' class compared to new segment class, which is expected, since 96.71\% instances belong to ``same segment'' categories. KNN achieves 22.40\%, 50.07\%, and 36.26\% higher precision, recall, and F1-measure, respectively, for lexical features; and 20.85\%, 33.96\%, and 27.47\% higher precision, recall, and F1-measure, respectively, for combined features compared to new segment classification. In contrast to new segment detection, RNN demonstrates the highest performance among all models. The impact of punctuation and topic model-based features is consistent with ``new segment'' classification. Results indicate that F1-measure increases by 0\%, and 0.1\% for SVM and KNN models and decreases by 0.1\% for NB. \\

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM, and KNN methods with different topic model-based features for segmentation of e-Coaching emails as a weighted average over ``same segment'' and ``new segment'' classification results, when all features are used together. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_weighted_avg}
  \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{L-LDA-based topic features}} & \multicolumn{3}{|c|}{\textbf{DL-LDA-based topic features}}  & \multicolumn{3}{|c|}{\textbf{LCA-based topic features}} \\\cline{2-10} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1}\\ \hline

@@

Table~\ref{tab:result_weighted_avg} summarizes the weighted average results of the models for segmentation of e-Coaching emails. Overall, SVM has the best performance across all topic model-based features in terms of all performance metrics, achieves 0.990 precision, recall and F1-measure when DL-LDA-based topic features are used. Similar to results in Tables~\ref{tab:result_base}, \ref{tab:result_boundary} and \ref{tab:result_not_boundary}, NB demonstrates the lowest performance among all methods while shows best performance with LCA-based topic features compared to other model-based topic features. Influence of the additional features is also consistent with the results in Tables~\ref{tab:result_boundary} and~\ref{tab:result_not_boundary}. Precision increases by 0\%, 1.43\%, and 0.2\%; recall increases by 0\%, 1.32\%, and 0.2\%; and F1-measure increases by 0\%, 1.32\%, and 0.3\% for NB, SVM, and KNN methods, respectively, when lexical feature is used in combination with punctuation and best model-based topic features.\\

>>>>

\author{Mehedi Hasan, BS$^{1}$\footnote[1]{Authors provided an equal contribution. \label{footnote1}}, Alexander Kotov, PhD$^{1}$\textsuperscript{\ref{footnote1}}, Sylvie Naar, PhD$^{2}$, Gwen L. Alexander, PhD$^{3}$, April Idalski Carcone, PhD$^{4}$}

@@

\noindent{\bf Abstract}
\textit{Communication science approaches to developing effective behavior interventions, such as motivational interviewing (MI), are limited by traditionally manual qualitative coding of communication exchanges, which is a very resource-intensive and time-consuming process. This study focuses on the analysis of e-Coaching sessions, behavior interventions that are delivered via email and grounded in the principles of MI. A critical step towards automated annotation of e-Coaching communication exchanges is segmentation of emails into textual fragments that correspond to MI behaviors. In this work, we formulate this task as a classification problem and propose lexical, punctuation and topic features to address it. We experimented both with traditional machine learning methods, such as Support Vector Machine (SVM), Naive Bayes and K-Nearest Neighbor (KNN) classifiers and recurrent neural networks (RNNs). Results indicate that SVM outperformed KNN and RNNs achieving 0.990 macro F1-score overall and 0.848 and 0.995 macro F1-score for detecting ``new segment'' and ``same segment'' classes, respectively.}

@@

Qualitative coding of motivational interviews with pre-defined codes has been traditionally performed manually by trained annotators, which is a tedious and resource-intensive process that involves several iterations of reading, comprehension and interpretation of interview transcripts. Rapidly developing computational technologies, specifically, machine learning methods, offer a unique opportunity to accelerate this process. In particular, machine learning methods have been successfully applied to a variety of analytical tasks involving textual data, such as classification\cite{nigam2000text} and sentiment analysis\cite{wang2012baselines}. In our previous work, we examined the utility of machine learning methods for automated annotation \cite{hasan2016study,kotov2015interpretable} and analysis \cite{hasan2018predicting} of in-person MI sessions. Specifically, we demonstrated that machine learning methods can be utilized for annotation of MI transcripts according to a simple communication code scheme with the accuracy comparable to human coders\cite{hasan2016study}. Experimental data utilized in these studies, however, were prepared by transcribing audio conversations, which were clearly segmented into utterances by a counselor, a patient and a caregiver.

@@

The goal of this research study is to assess the applicability of machine learning methods for automated segmentation of e-Coaching emails into textual fragments corresponding to individual behaviors, which is the first step of the coding process of e-Coaching communications. In particular, we introduced lexical, topic and punctuation features and experimented with both traditional supervised machine learning methods, such as Support Vector Machine (SVM), Naive Bayes (NB) and K-Nearest Neighbor (KNN) classifiers and deep learning methods, such as Long Short Term Memory (LSTM) and Gated Recurrent Unit (GRU), to find the best performing method and feature combination.

Relevant previous work in the biomedical domain primarily focused on segmentation of text into sections and headers\cite{apostolova2009automatic,denny2009evaluation,tepper2012statistical,cho2002text} or sentence boundary detection\cite{griffis2016quantitative,kreuzthaler2015detection,treviso2016sentence}. Apostolova et al.\cite{apostolova2009automatic} applied SVM along with word-vector cosine similarity metric combined with several heuristics to segment clinical reports into sections, such as demographics, history, procedure, finding and impression. After identification of each line in the document, Tepper et al. \cite{tepper2012statistical} trained Maximum Entropy models for section classification. Denny et al.\cite{denny2009evaluation} proposed a SecTag algorithm, which combined natural language processing techniques, terminology-based rules and Naive Bayes classifier to identify the sections and headers that achieved 99\% recall with 95.6\% precision. On the other hand, SVM based on prosodic and part of speech features \cite{kreuzthaler2015detection} and recurrent convolutional neural networks using word embeddings \cite{griffis2016quantitative} were utilized for detecting sentence boundaries. Segmentation of e-Coaching emails is different from a traditional shallow discourse analysis of conversations \cite{galley2003discourse} in that the focus is on segmentation, rather than on determining the types of transitions between the utterances or assigning utterances to speakers.

@@

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from 49 e-coaching sessions, which include a total of 3,138 segmented and annotated MI behaviors. Each session represents an MI intervention delivered via email. During pre-processing, we removed all non-ASCII characters and applied stemming to normalize morphological variants of related concepts, such as ``eating'', ``eats'' and ``eat''. We formulate the segmentation task as a binary classification problem, as illustrated in Figure~\ref{fig:classifier} and experimented both with traditional machine learning methods, such as Support Vector Machine (SVM), Naive Bayes (NB) and K-Nearest Neighbor (KNN) classifiers and recurrent neural networks (RNNs), such as Long Short Term Memory (LSTM) and Gated Recurrent Unit (GRU). For NB, SVM and KNN models, e-Coaching email exchanges are partitioned into adjacent word pairs. Each pair is classified into either ``new segment'' or ``same segment'' class. The gold standard is manually segmented where adjacent word pairs are classified into ``new segment'' class. If all adjacent word pairs of a block of text are classified into ``same segment'' class, the entire block is treated as one textual fragment corresponding to a single MI behavior. In total, we obtained 95,421 word pairs, which include 3,138 ``new segment'' and 92,283 ``same segment'' instances. For RNNs, a block of text was taken as input sequence, such that one-hot encodings of each word or punctuation marks in a block were used as input into an RNN and binary labels (1 or 0) corresponding to ``new segment'' and ``same segment'' classification decision were considered as the output of RNN at each step. In the gold standard, words within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1.

@@

where $p(l)$ is the prior probabilities of label $l$ in the training set and $p(w)$ is the prior probability of word $w$ in the language model estimated from the training set using maximum likelihood. We use the distribution $p(l|w)$ as the topic feature vector, which shows how indicative each word $w$ is for a particular behavior code $l$. Topic features allow to abstract away from individual words and detect segment boundaries by indirectly capturing transitions between behavior codes when only observing a pair of words that are specific to different behavior codes. To find the optimal topic feature, we also derived the same topic feature $p(l|w)$ by using two recently proposed probabilistic generative latent variable models: Discriminative Labeled Latent Dirichlet Allocation (DL-LDA) and Latent Class Allocation (LCA)\cite{kotov2015interpretable}. Similar as L-LDA, DL-LDA-based topic feature is extracted by first deriving a label-specific language model $p(w|l)$ for each label $l$ from the word distributions $p(w|l,z)$ for topic $z$ specific to label $l$ by using the Eq.~\ref{eq:pwl} and \ref{eq:plw}. On the other hand, LCA directly provides label-specific multinomials $p(w|l)$, which can be used to derive topic feature by using Eq.~\ref{eq:plw}. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;', `-'\} between a pair of words, are also employed as a feature since punctuation marks designate the boundary of a sentence, clause and phrase and often also correspond to segment boundary.

@@

\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\label{eq:resetgru}
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\label{eq:candidategru}
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
In Eq.~\ref{eq:firstgru}-\ref{eq:lastgru}, $\sigma$ corresponds to sigmoid function and $\odot$ designates an element-wise product. The update gate $z_t$ and reset gate $r_t$ at time step $t$ are computed by the Eq.~(\ref{eq:firstgru}) and~(\ref{eq:resetgru}), where $W_z$, $W_r$, $W_h$, $U_z$, $U_r$, $U_h$ are the weight matrices and $b_z$, $b_h$ and $b_r$ are bias vectors. The activation $h_t$ of the GRU at time $t$ is a linear combination of previous activation $h_{t-1}$ and the candidate activation $\tilde h_t$, which is represented by Eq.~(\ref{eq:lastgru}) and~(\ref{eq:candidategru}). Our GRU model includes one hidden layer, output layer and input layer. We reset our model state after feeding each input sequence where input was given as one-hot encoding of word vector. Since LSTM and GRU use one-hot vector as input, results for these models are reported only when lexical and punctuation features are used. We experimentally determined that the best performance is achieved when the number of hidden units is 25, batch size is 1 and Adam\cite{kingma2014adam} is used for optimization.

\subsection*{\textit{Evaluation metrics}}
We report standard metrics for experiments (precision, recall and F1-measure) to evaluate the performance of binary classifiers\cite{aas1999text}. However, accuracy is not reported as a performance metric because accuracy is highly sensitive to the prior class probabilities and does not fully describe the actual difficulty of the decision problem for an unbalanced dataset. The results are reported based on 5 folds cross-validation and weighted macro-averaging over the folds. We also estimate the area under the receiving operating characteristics (ROC) curve\cite{kumar2011receiver} (AUC) metric due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets \cite{hu2015kernelized}.

\section*{Results}
Our experimental results have two important directions. First, results are reported with respect to ``new segment'' and ``same segment'' classes as well as their weighted average in Table~\ref{tab:result_base}. Second, classification performance of different machine learning methods are outlined in Tables~\ref{tab:result_boundary}, \ref{tab:result_not_boundary} and \ref{tab:result_weighted_avg} when topic features are extracted with different probabilistic generative latent variable models.\\

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM and KNN methods for identification of ``new segment'' and ``same segment'' classes as well as their weighted average. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_base}
  \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{new segment}} & \multicolumn{3}{|c|}{\textbf{same segment}}  & \multicolumn{3}{|c|}{\textbf{overall}} \\\cline{2-10}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1}\\ \hline    
 \multicolumn{10}{|c|}{lexical features only} \\ \hline   
 NB & 0.594 & 0.662 & 0.626 & 0.988 & 0.985 & 0.987 & 0.975 & 0.974 & 0.975  \\ \hline
 SVM & 0.762 & \textbf{0.673} & 0.715 & \textbf{0.989} & 0.993 & 0.991  & 0.981 & 0.982 & 0.982 \\ \hline
 KNN & \textbf{0.808} & 0.663 & \textbf{0.728} & \textbf{0.989} & \textbf{0.995} & \textbf{0.992} & \textbf{0.983} & \textbf{0.984} & \textbf{0.983} \\ \hline
 \multicolumn{10}{|c|}{lexical and puctuation features} \\ \hline
 LSTM & \textbf{0.800} & 0.646 & 0.714  & 0.993 & \textbf{0.994} & \textbf{0.994} & \textbf{0.986} & 0.983 & 0.984 \\ \hline
 GRU & \textbf{0.800} & \textbf{0.715} & \textbf{0.741} & \textbf{0.994} & \textbf{0.994} & \textbf{0.994} & \textbf{0.986} & \textbf{0.985} & \textbf{0.986} \\ \hline
  \end{tabular}
\end{table}

As follows from Table~\ref{tab:result_base}, KNN outperforms all other models in terms of precision and F1-measure achieving 0.808 precision with 0.728 F1-measure for new segment detection. KNN also shows superior performance in all performance metrics for ``same segment'' class and weighted average over ``new segment'' and ``same segment'' classes. NB demonstrates the lowest performance among all models in terms of all performance metrics. On the other hand, SVM has the highest recall of 0.673 when only lexical features are used to identify ``new segment''. Results indicate that performance of classifiers is remarkably higher for ``same segment'' class compared to ``new segment'' class, which is expected since 96.71\% instances belong to ``same segment'' categories. For example, KNN achieves 22.40\%, 50.07\% and 36.26\% higher precision, recall and F1-measure, respectively, in same segment identification compared to new segment detection. When lexical features are used in combination with punctuation mark, GRU demonstrates the best performance in all cases showing 10.68\% higher recall and 3.78\%  higher F1-measure than LSTM for new segment identification, 0.1\% higher precision than LSTM for same segment identification and 0.2\% higher recall and 0.2\%  higher F1-measure than LSTM for overall classification. \\

@@

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM and KNN methods with different topic model-based features for identification of ``new segment'' class in e-Coaching emails, when all features are used together. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_boundary}
  \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{L-LDA-based topic features}} & \multicolumn{3}{|c|}{\textbf{DL-LDA-based topic features}}  & \multicolumn{3}{|c|}{\textbf{LCA-based topic features}} \\\cline{2-10} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1}\\ \hline

@@

Table~\ref{tab:result_boundary} summarizes the performance of NB, SVM and KNN models for detecting ``new segment'' class. We observed that classification performance significantly improved for ``new segment'' class when lexical features are used in combination with punctuation and model-based topic features. Precision increases by 4.99\% and 0\%, 11.29\% and 0.74\% and 9.71\% and 0.74\% for SVM and KNN; recall increases by 18.72\% and 17.06\%, 26.15\% and 15.64\% and 24.67\% and 16.90\% for SVM and KNN; and F1-measure increases by 11.75\% and 6.18\%, 18.60\% and 5.91\% and 17.06\% and 6.46\% for SVM and KNN when L-LDA, DL-LDA and LCA-based topic features are used in conjunction combination with lexical and punctuation features, respectively. On the other hand, NB has very negligible impact of punctuation and model-based topic features. In contrast to results in Table~\ref{tab:result_base}, SVM demonstrates the best performance among all models achieving 0.848, 0.849 and 0.848 for precision, recall and F1-measure, respectively, when DL-LDA-based topic features are used in conjunction with lexical and punctuation features. NB and KNN provide their best performance when LCA-based topic features are used, achieving 0.626 and 0.775 F1-measure, respectively.\\

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM and KNN methods with different topic model-based features for identification of ``same segment'' class in e-Coaching emails, when all features are used together. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_not_boundary}
  \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{L-LDA-based topic features}} & \multicolumn{3}{|c|}{\textbf{DL-LDA-based topic features}}  & \multicolumn{3}{|c|}{\textbf{LCA-based topic features}} \\\cline{2-10} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1}\\ \hline

@@

Similar to baseline performance, classification performance is remarkably higher for same segment class compared to new segment class (Table~\ref{tab:result_not_boundary}). The impact of punctuation and topic model-based features is also consistent with ``new segment'' classification while improvements are not as significant as ``new segment'' class. We observed that precision increases by 0.4\% and 0.2\%, 0.6\% and 0.2\% and 0.6\% and 0.2\% for SVM and KNN; recall increases by 0\% and -0.1\%, 0.2\% and -0.1\% and 0.1\% and -0.1\% for SVM and KNN; and F1-measure increases by 0.2\% and 0.1\%, 0.4\% and 0.1\% and 0.3\% and 0.1\% for SVM and KNN when L-LDA, DL-LDA and LCA-based topic features are used in conjunction combination with lexical and punctuation features, respectively. Similar to results in Table \ref{tab:result_boundary}, NB and KNN demonstrate the lowest and moderate performance on the task of identifying same segment, respectively. \\

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM and KNN methods with different topic model-based features for segmentation of e-Coaching emails as a weighted average over ``same segment'' and ``new segment'' classification results, when all features are used together. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_weighted_avg}
  \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{L-LDA-based topic features}} & \multicolumn{3}{|c|}{\textbf{DL-LDA-based topic features}}  & \multicolumn{3}{|c|}{\textbf{LCA-based topic features}} \\\cline{2-10} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1}\\ \hline

@@

Table~\ref{tab:result_weighted_avg} summarizes the weighted average results of the models for segmentation of e-Coaching emails. Overall, SVM has the best performance across all topic model-based features in terms of all performance metrics, achieves 0.990 precision, recall and F1-measure when DL-LDA-based topic features are used. Similar to results in Tables~\ref{tab:result_base}, \ref{tab:result_boundary} and \ref{tab:result_not_boundary}, NB demonstrates the lowest performance among all methods while shows best performance with LCA-based topic features compared to other model-based topic features. Influence of the additional features is also consistent with the results in Tables~\ref{tab:result_boundary} and~\ref{tab:result_not_boundary}. Precision increases by 0.9\% and 0.2\%; recall increases by 0.8\% and 0.2\%; and F1-measure increases by 0.8\% and 0.3\% for SVM and KNN methods, respectively, when lexical feature is used in combination with punctuation and best model-based topic features.\\
